Analyzing Information Transfer from Different Data Sources via Precise Bias-variance Asymptotics

### 2021-07-07 revision_journal/results.tex
Updates from Overleaf

In case (i), we provide the exact asymptotics for both the bias and variance terms, together with almost sharp convergence rates. We will use these results to illustrate the effects of the \emph{bias-variance tradeoff} and \emph{varying sample sizes}. In case (ii), we will prove an exact asymptotic variance limit, and employ the result to show the intricate effect of the covariate shift. On the other hand, in case (ii), we will provide an estimate on the bias term, which becomes exact only when $n_1\gg p$. %Here case (ii) is actually more general than (i)

@@

%By solving $B$ in equation \eqref{eq_mtl}, we obtain the estimator for task two as follows:
%\begin{align}
%	\hat{\beta}_2^{\MTL} = {\hat{\Sigma}}^{-1} ({X^{(1)}}^{\top} Y^{(1)} + {X^{(2)}}^{\top} Y^{(2)}), \text{ where }
%	\hat{\Sigma} = {X^{(1)}}^{\top} X^{(1)} + {X^{(2)}}^{\top} X^{(2)}. \label{def hatsig}
%\end{align}
First, we consider the variance. Recall from equation \eqref{Lvar} that the variance equation is equal to $\sigma^2 \cdot \bigtr{\Sigma^{(2)} (\hat\Sigma(a))^{-1}}$.
In particular, the matrix $\hat{\Sigma}$ adds up both tasks' sample covariance matrices.
Thus, the expectation of $\hat{\Sigma}(a)$ is equal to a mixture of both tasks' population covariance matrices, with mixing proportions determined by their sample sizes.
%To derive the bias and variance equation, we consider the expected loss conditional on the covariates as follows (the empirical loss is close to this expectation as will be shown in equation \eqref{claim_largedev2}):
 %similar to Claim \ref{claim_pred_err}
%\begin{align}
%	 \exarg{\cE}{L(\hat{\beta}_2^{\MTL}) \mid X^{(1)}, X^{(2)}}
%	=& \bignorm{{\Sigma^{(2)}}^{1/2} \hat{\Sigma}^{-1} {X^{(1)}}^{\top} X^{(1)} (\beta^{(1)} - \beta^{(2)})}^2 \label{eq_bias_2task} \\
%	& + \sigma^2 \bigtr{\Sigma^{(2)}\hat{\Sigma}^{-1}}. \label{eq_variance_2task}
%\end{align}
%Equations \eqref{eq_bias_2task} and \eqref{eq_variance_2task} correspond to the bias and variance of HPS for two tasks, respectively.
%Our key result characterizes the asymptotic limit of the inverse of the sum of two arbitrarily different sample covariance matrices.
Intuitively, the spectrum of $\hat{\Sigma}(a)^{-1}$ now not only depends on the sample sizes of both tasks, but also depends on how well ``aligned'' $\Sigma^{(1)}$ and $\Sigma^{(2)}$ are.
%However, capturing this intuition quantitatively turns out to be technically challenging.
%The main technical challenge of our result deals with the ``covariate shift'' between tasks one and two.
To capture this alignment quantitatively, we introduce the covariate shift matrix %(rescaled by $a$)
$$ M \define (\Sigma^{(1)})^{\frac 1 2}(\Sigma^{(2)})^{-\frac 1 2}.$$
%Let $U\Lambda V^\top$ denote the SVD of $M$ and
Let $\lambda_1 \ge \lambda_2 \ge \dots\ge \lambda_p $ be the singular values of $M$ in descending order. %By \eqref{assm3}, these singular values satisfy that
%\begin{equation}\label{assm32}
%|a| \tau \le  \lambda_p(a) \le \cdots \le \lambda_2(a)\le \lambda_1(a) \le |a|\tau^{-1}. %, \quad \max\left\{\pi_A^{(n)}([0,\tau]), \pi_B^{(n)}([0,\tau])\right\} \le 1 - \tau .
%\end{equation}
%for some small constant $0<\tau<1$.
Our first main result is the following theorem on the variance limit, which characterizes the exact dependence of $L_{\var}(a)$ on the singular values of $M$.

@@

Equation \eqref{lem_cov_shift_eq} thus characterizes the variance of the HPS estimator, up to an error term on the order of $O\big(\frac{\sigma^2}{p^{-1/2}}\big)$ (recall that $\varphi > 4$, $c$ is an arbitrarily small constant, $\frac{p}{n_1+n_2} \in (2 {\tau}, \frac 2 {\tau})$).
Our result can be viewed as a generalization of the classical result in Lemma \ref{fact_tr} to the \textit{covariate shift} setting.
In particular, Lemma \ref{fact_tr} corresponds to a setting where the singular values $\lambda_1,\dots,\lambda_p$ are all equal to one, and $a = 1$. Then, the second part of equation \eqref{eq_a12extra} simplifies to
\[ \alpha_1 + \frac{p}{n_1 + n_2}\cdot \frac{\alpha_1}{\alpha_1 + \alpha_2} = \frac{n_1}{n_1 + n_2}. \]
By solving the above together with the first part of equation \eqref{eq_a12extra}, we obtain that
$$\alpha_1 = \frac{n_1}{n_1+n_2},\quad \alpha_2 = \frac{n_2-p}{n_1+ n_2}.$$
Plugging the above solutions back to the variance limit, we thus find that the limit is equal to
\begin{align}
    \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}} &= \frac{\sigma^2}{n_1 + n_2} \frac{p}{\alpha_1 + \alpha_2} \tag{since the singular values of $M$ are equal to $1$} \nonumber \\
    &= \frac{\sigma^2}{n_1 + n_2} \cdot\frac{p (n_1 + n_2)}{n_1 + n_2 - p} \tag{applying $\alpha_1$ and $\alpha_2$ from above} \nonumber \\
    &= \frac{\sigma^2 p}{n_1 + n_2 - p}, \label{g_id}
\end{align}
which is precisely the limit in Lemma \ref{fact_tr}.
Additionally, the convergence rate in equation \eqref{lem_cov_shift_eq} matches the one in Lemma \ref{fact_tr} for this special case.
%With Assumption \ref{assm_big1},
%$$ \frac{\sigma^2}{n_1+n_2}\bigtr{  \frac{1}{a_1 M(a)^\top M(a) + a_2  }  } \sim \frac{p \sigma^2}{n_1+ n_2}.$$
%Hence the right-hand side of \eqref{lem_cov_shift_eq} is much smaller than this main term by a factor of $p^{-1/2} (n_1+n_2)^{-1/2+2/\varphi + c}$, which we believe to be sharp up to the $(n_1+n_2)^c$ factor. Lemma \ref{fact_tr} can be also regarded as a special case of Theorem \ref{thm_main_RMT}. %also extends  to the inverse of the sum of two sample covariance matrices.
%To see this, when $M(a)=0$,
The proof of Theorem \ref{thm_main_RMT}, which is based on recent developments in random matrix theory \cite{Anisotropic}, can be found in Section \ref{appendix RMT}.
%For general $A_1,A_2$ that are not equal to one, we can still apply our result by rescaling $X^{(1)}$ and $M$ with $A_1 / A_2$.
%We defer a proof sketch of Theorem \ref{thm_main_RMT} until the end of the section.
%This amounts to replacing $M$ with $\frac{A_1}{A_2}M$ in Theorem \ref{thm_main_RMT}.

@@

\begin{corollary}[Asymptotics for HPS under covariate shift]\label{cor_hps_cov}
    Under Assumption \ref{assm_big1}, suppose further that $\beta_1 = \beta_2$ and $\norm{\beta_1}^2_2 \ge \frac{\sigma^2}{p^{1/2 - c_0}}$ for a constant $c_0 > 0$.
    Then, for any constant $c > 0$, the following estimate on the excess risk of the HPS estimator holds w.h.p.
    \begin{align}
        \bigabs{L(\hat{\beta}_2^{\MTL}) - \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac{1}{\alpha_1 M^{\top} M + \alpha_2}}}
        \le
        \OO\bigbrace{\frac{(n_1+n_2)^{\frac 2{\varphi} - \frac 1 2 + c}}{p^{1/2}}\cdot\frac{p \sigma^2}{n_1+ n_2}
        + p^{-c_0 + 2c}},
    \end{align}
    where $\alpha_1$ and $\alpha_2$ are the solutions of equation \eqref{eq_a12extra} after taking $a = 1$.
\end{corollary}

\begin{figure}[!t]
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=0.9\textwidth]{figures/covariate_shift.eps}
		\caption{HPS estimator under covariate shift}
		\label{fig_sec3_covariate}
	\end{subfigure}
	\caption{The dimension $p = 200$, $n_2 = 800$, and the noise variance $\sigma^2 = 1/4$.}
\end{figure}

\subsubsection{Illustrative examples}

Next, we illustrate the result of Corollary \ref{cor_hps_cov} in several examples.

\begin{example}[Does covariate shift help or hurt information transfer?]
    %However, we now use an example to show that the above observation fails when $n_1$ is comparable to $n_2$.
    Our first example illustrates that the effect of covariate shift depends on the sample sizes of each dataset in an intricate manner.
    While the folklore belief is that transferring from a covariate-shifted distribution performs worse than an identical distribution, we show that, surprisingly, the former can sometimes outperform the latter.

@@

\begin{proof}
    We explain why the above dichotomy happens. We can write $g(M)$ for any $M \in \cS$ as
    $$g(M)=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2}\left( \frac{1}{\lambda_i^{2} a_1 + a_2} + \frac1{\lambda_i^{-2} a_1 + a_2} \right).$$
    When $M=\id_{p\times p}$, by the first equation of \eqref{eq_a12extra}, we have
    $$g(\id_{p\times p})=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2} \frac{2}{1-\gamma},$$
    where we abbreviate $\gamma:=p/(n_1+n_2)$. Then using $\alpha_1 + \alpha_2 = 1-\gamma$, through a direct calculation we find that %we can calculate that
    \begin{align*}
        g(M) - g(\id_{p\times p})%&= \frac{\sigma^2 \gamma}{2(1-\gamma)} (\lambda^2-1)a_1\cdot \bigbrace{  \frac{1}{ -a_1(\lambda^2-1)+(1-\gamma)\lambda^2 } - \frac{1}{a_1(\lambda^2-1) + (1-\gamma)}} \\
        &= \frac{\sigma^2 }{ n_1+n_2-p} \sum_{i=1}^{p/2} \frac{(\lambda_i^2-1)^2 \alpha_1 \left[ \alpha_1 - \alpha_2\right] }{[\alpha_1 + \lambda_i^2 \alpha_2][\lambda_i^2 \alpha_1 + \alpha_2]} .
    \end{align*}
    %in this example is equal to
    %$\frac{p}{2(n_1 + n_2)} f(\lambda)$, where
    %\[ f(\lambda) = {(\lambda^{-2} a_1 + a_2)^{-1} + (\lambda^2 a_1 + a_2)^{-1}}. \]
    %Using the fact that $a_1 + a_2 = 1 - \frac{p}{n_1 + n_2}$, we can verify
    %\begin{align*}
    %	f(\lambda) - f(1) &= \left(\lambda^2 a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
    %	&+ \left(\lambda^{-2} a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
    %	&- \frac{2(n_1 + n_2)}{n_1 + n_2 - p} \\
    %	&= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
    %\end{align*}
    %\begin{align*}
    %	f(\lambda) - f(1) &= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
    %\end{align*}
    %where $g(\lambda, a_1) \ge 0$.
    %and can be derived from algebraic calculations (details omitted).
    We claim that $\alpha_1 > \alpha_2$ if and only if $n_1 > n_2$, which then explains the dichotomy.
    In fact, if $\alpha_1 > \alpha_2$, then the first part of equation \eqref{eq_a12extra} gives that $\alpha_1 > (1-\gamma)/2$.
    The second part of equation \eqref{eq_a12extra} gives that
    \begin{align*}
     \frac{n_1}{n_1 + n_2} &> \alpha_1 + \frac{1}{n_1+n_2} \sum_{i=1}^{p/2}\left(\frac{\lambda_i^2}{\lambda_i^2+1}+\frac{\lambda_i^{-2}}{\lambda_i^{-2}+1}\right) = \frac{1-\gamma}{2}+\frac{\gamma}{2}=\frac{1}{2}.
    \end{align*}
    %\begin{align*}
    % \frac{n_1}{n_1 + n_2} &=a_1 + \frac1{n_1 + n_2}\cdot \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2}} \\
    %	&> a_1 + \frac{p}{2(n_1+n_2)} \left(\frac{\lambda^2}{\lambda^2+1}+\frac{\lambda^{-2}}{\lambda^{-2}+1}\right) =\frac{1}{2}.
    %\end{align*}
    This is equivalent to $n_1>n_2$. The other direction follows from a similar argument. % Similarly, if $a_1<a_2$, equations  \eqref{eq_a12extra000} and  \eqref{eq_a12extra} give that $a_1 < \frac{n_1 + n_2-p}{2 (n_1 + n_2)}$ and $n_1<n_2$. Thus, we conclude that $f(\lambda) \ge f(1)$ if and only if $n_1 \ge n_2$.
    %\end{example}
\end{proof}

Figure \ref{fig_sec3_covariate} illustrates a special case where $\lambda_1 = \cdots = \lambda_{\floor{p /2}} = \lambda > 0$ for $M^{\top}M$ and the rest of the eigenvalues are all equal to $1 / \lambda$.
Thus, $\lambda > 0$ captures the degree of covariate shift: higher $\lambda$ implies worse covariate shift.
We observe that our theoretical estimates using Corollary \ref{cor_hps_cov} matches the empirical risks incredibly accurately.
As a result, we indeed observe the dichotomy shown in Claim \ref{claim_dichotomy}.
\end{example}

%\paragraph{Illustrative examples.}
%Hence we take $a=1$ in \eqref{lem_cov_shift_eq}, and study the asymptotic limit
%$$\ell_{\var}(M):= \frac{\sigma^2}{n_1+n_2}\bigtr{  \frac{1}{a_1(M) \cdot M^\top M + a_2(M)  }  }, $$
%where $ M=(\Sigma^{(1)})^{1/2}(\Sigma^{(2)})^{-1/2}$, and $a_1(M)$ and $a_2(M)$ are defined through \eqref{eq_a12extra}.
\begin{example}
Our second example illustrates that the effect of covariate shift worsens as the sample size of task one increases.
We consider a set of covariate shift matrices $M^{\top} M$ whose determinant are all equal to one, and the eigenvalues are bounded between $\tau^2$ and $1 / \tau^2$.
Let $\cS$ denote the set of $M\in\real^{p\times p}$ such that $M^{\top} M$ satisfies both conditions.
%\begin{align*}
%		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i = \mu^p, \tau \le \lambda_p \le \lambda_1 \le \tau^{-1}\right.}.
%\end{align*}
\begin{claim}[Identical covariance provides approximately optimal transfer under imbalanced dataset sizes]\label{prop_covariate}
%	Assume that $\Psi(\beta_1, \beta_2) = 0$ and $\rho_1, \rho_2>1$.
%	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} \in\cS_{\mu}$.
    Recall that $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
	We have the following relation between $g(\id_{p \times p})$ and the rest of $g(M)$ for any $M \in\cS$:
	\begin{align} g(\id_{p\times p}) \le \bigbrace{1+ {\frac{n_2}{\tau^2\cdot n_1}  }} g(M), \text{ for any } M \in \cS. \label{eq_claim_id}
	\end{align}
\end{claim}
\begin{proof}
    We can write down the trace of $(\alpha_1 M^{\top} M + \alpha_2)^{-1}$ using the eigenvalues of $M^{\top} M$ as follows:
    \begin{align*}
        (\alpha_1 M^{\top} M + \alpha_2)^{-1} &= \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2} \\
        &\ge \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2 \cdot {\lambda_i^2} / {\tau^2}} \tag{since $\lambda_i \le \tau$, for any $i$} \\
        &= \frac 1 {\alpha_1 + \alpha_2 / \tau^2} \sum_{i=1}^ p \frac 1 {\lambda_i^2} \\
        &\ge \frac 1 {\alpha_1 + \alpha_2 /\tau^2} p\cdot \bigbrace{\prod_{i=1}^p \frac 1 {\lambda_i^2}}^{1/ p} \tag{by the AM-GM inequality} \\
        &= \frac{p}{\alpha_1 + \alpha_2 / \tau^2} \tag{since $\prod_{i=1}^p \lambda_i^2 = \det(M^{\top} M) = 1$}
    \end{align*}
    Next, recall from equation \eqref{g_id} that
    $g(\id_{p\times p}) = \frac{\sigma^2 p}{n_1 + n_2 - p}$.
    Thus, equation \eqref{eq_claim_id} follows if we show that (after rearranging terms)
    \begin{align}
        \frac{n_1 + n_2}{n_1 + n_2 - p} \cdot \frac 1 {\alpha_1 + \frac{\alpha_2}{\tau^2}} \le 1 + \frac{n_2}{\tau^2 \cdot n_1}. \label{eq_claim_id_step}
    \end{align}
    From the second part of equation \eqref{eq_a12extra}, we have that
    \begin{align*}
        \frac{n_1} {n_1 + n_2} &= \alpha_1 + \frac 1 {n_1 + n_2} \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2\alpha_1}{\lambda_i^2\alpha_1 + \alpha_2}} \\
        &< \alpha_1 + \frac p {n_1 + n_2}.
    \end{align*}
    Thus, $\alpha_1 > \frac{n_1 - p}{n_1 + n_2}$, which implies that $\alpha_2 < \frac{n_2}{n_1 + n_2}$.
    Hence, $\alpha_1 + \frac{\alpha_2}{\tau^2} = 1 - \frac{p}{n_1 + n_2} - \alpha_2 + \frac{\alpha_2}{\tau^2} \le 1 - \frac{p}{n_1 + n_2} + \frac{n_2}{n_1 + n_2} \frac 1 {\tau^2}$, which implies
    equation \eqref{eq_claim_id_step} by straightforward calculations.
    Thus, the proof is complete.
\end{proof}
\todo{draw a connection to \citet{david2010impossibility}}
\end{example}

\subsubsection{Comparing HPS to other estimators}

>>>>

In case (i), we provide the exact asymptotics for both the bias and variance terms, together with almost sharp convergence rates. We will use these results to illustrate the effects of the \emph{bias-variance tradeoff} and \emph{varying sample sizes}. In case (ii), we will prove an exact asymptotic variance limit, and employ the result to show the intricate effect of the covariate shift. On the other hand, in case (ii), we will provide an estimate on the bias term, which becomes exact only when $n_1\gg p$.

@@

First, we consider the variance. Recall from equation \eqref{Lvar} that the variance equation is equal to $\sigma^2 \cdot \bigtr{\Sigma^{(2)} (\hat\Sigma(a))^{-1}}$.
In particular, the matrix $\hat{\Sigma}$ adds up both tasks' sample covariance matrices.
Thus, the expectation of $\hat{\Sigma}(a)$ is equal to a mixture of both tasks' population covariance matrices, with mixing proportions determined by their sample sizes.
Intuitively, the spectrum of $\hat{\Sigma}(a)^{-1}$ now not only depends on the sample sizes of both tasks, but also depends on how well ``aligned'' $\Sigma^{(1)}$ and $\Sigma^{(2)}$ are.
To capture this alignment quantitatively, we introduce the covariate shift matrix %(rescaled by $a$)
$$ M \define (\Sigma^{(1)})^{\frac 1 2}(\Sigma^{(2)})^{-\frac 1 2}.$$
Let $\lambda_1 \ge \lambda_2 \ge \dots\ge \lambda_p $ be the singular values of $M$ in descending order.
Our first main result is the following theorem on the variance limit, which characterizes the exact dependence of $L_{\var}(a)$ on the singular values of $M$.

@@

Equation \eqref{lem_cov_shift_eq} thus characterizes the variance of the HPS estimator, up to an error term on the order of $O\big(\frac{\sigma^2}{p^{-1/2}}\big)$ (recall that $\varphi > 4$, $c$ is an arbitrarily small constant, $\frac{p}{n_1+n_2} \in (2 {\tau}, \frac 2 {\tau})$).
Our result can be viewed as a generalization of the classical result in Lemma \ref{fact_tr} to the \textit{covariate shift} setting.
In particular, Lemma \ref{fact_tr} corresponds to a setting where the singular values $\lambda_1,\dots,\lambda_p$ are all equal to one, and $a = 1$. Then, the second part of equation \eqref{eq_a12extra} simplifies to
\[ \alpha_1 + \frac{p}{n_1 + n_2}\cdot \frac{\alpha_1}{\alpha_1 + \alpha_2} = \frac{n_1}{n_1 + n_2}. \]
By solving the above together with the first part of equation \eqref{eq_a12extra}, we obtain that
$$\alpha_1 = \frac{n_1}{n_1+n_2},\quad \alpha_2 = \frac{n_2-p}{n_1+ n_2}.$$
Plugging the above solutions back to the variance limit, we thus find that the limit is equal to
\begin{align}
    \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}} &= \frac{\sigma^2}{n_1 + n_2} \frac{p}{\alpha_1 + \alpha_2} \tag{since the singular values of $M$ are equal to $1$} \nonumber \\
    &= \frac{\sigma^2}{n_1 + n_2} \cdot\frac{p (n_1 + n_2)}{n_1 + n_2 - p} \tag{applying $\alpha_1$ and $\alpha_2$ from above} \nonumber \\
    &= \frac{\sigma^2 p}{n_1 + n_2 - p}, \label{g_id}
\end{align}
which is precisely the limit in Lemma \ref{fact_tr}.
Additionally, the convergence rate in equation \eqref{lem_cov_shift_eq} matches the one in Lemma \ref{fact_tr} for this special case.
%With Assumption \ref{assm_big1},
%$$ \frac{\sigma^2}{n_1+n_2}\bigtr{  \frac{1}{a_1 M(a)^\top M(a) + a_2  }  } \sim \frac{p \sigma^2}{n_1+ n_2}.$$
%Hence the right-hand side of \eqref{lem_cov_shift_eq} is much smaller than this main term by a factor of $p^{-1/2} (n_1+n_2)^{-1/2+2/\varphi + c}$, which we believe to be sharp up to the $(n_1+n_2)^c$ factor. Lemma \ref{fact_tr} can be also regarded as a special case of Theorem \ref{thm_main_RMT}. %also extends  to the inverse of the sum of two sample covariance matrices.
The proof of Theorem \ref{thm_main_RMT}, which is based on recent developments in random matrix theory \cite{Anisotropic}, can be found in Section \ref{appendix RMT}.

@@

\begin{corollary}[Excess risk of HPS under covariate shift]\label{cor_hps_cov}
    Under Assumption \ref{assm_big1}, suppose further that $\beta_1 = \beta_2$ and $\norm{\beta_1}^2_2 \ge \frac{\sigma^2}{p^{1/2 - c_0}}$ for a constant $c_0 > 0$.
    Then, for any constant $c > 0$, the following estimate on the excess risk of the HPS estimator holds w.h.p.
    \begin{align}
        \bigabs{L(\hat{\beta}_2^{\MTL}) - \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac{1}{\alpha_1 M^{\top} M + \alpha_2}}}
        \le
        \OO\bigbrace{\frac{(n_1+n_2)^{\frac 2{\varphi} - \frac 1 2 + c}}{p^{1/2}}\cdot\frac{p \sigma^2}{n_1+ n_2}
        + p^{-c_0 + 2c}},
    \end{align}
    where $\alpha_1$ and $\alpha_2$ are the solutions of equation \eqref{eq_a12extra} after taking $a = 1$.
\end{corollary}

\begin{figure}[!t]
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=0.9\textwidth]{figures/verify_covariate_shift.eps}
		\caption{Verify variance asymptotics in finite dimensions}
		\label{fig_sec3_covariate}
	\end{subfigure}
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=0.9\textwidth]{figures/covariate_shift.eps}
		\caption{HPS estimator under covariate shift}
		\label{fig_sec3_covariate}
	\end{subfigure}
	\caption{Figure \ref{fig_sec3_covariate}: we verify that Theorem \ref{thm_main_RMT} provides incredibly accurate estimates of the empirical variance under various finite sample sizes and dimensions.
	Figure \ref{fig_sec3_covariate}: We illustrate an intriguing dichotomy between covariate shift and sample size ratio: The dimension $p = 200$, $n_2 = 800$, and the noise variance $\sigma^2 = 1/4$.}
\end{figure}

\paragraph{Illustrative examples.}
Next, we illustrate the result of Corollary \ref{cor_hps_cov} in several examples.

\begin{example}[Does covariate shift help or hurt information transfer?]
    Our first example illustrates that the effect of covariate shift depends on the sample sizes of each dataset in an intricate manner.
    While the folklore belief is that transferring from a covariate-shifted distribution performs worse than an identical distribution, we show that, surprisingly, the former can sometimes outperform the latter.

@@

\begin{proof}
    We explain why the above dichotomy happens. We can write $g(M)$ for any $M \in \cS$ as
    $$g(M)=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2}\left( \frac{1}{\lambda_i^{2} a_1 + a_2} + \frac1{\lambda_i^{-2} a_1 + a_2} \right).$$
    When $M=\id_{p\times p}$, by the first equation of \eqref{eq_a12extra}, we have
    $$g(\id_{p\times p})=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2} \frac{2}{1-\gamma},$$
    where we abbreviate $\gamma:=p/(n_1+n_2)$. Then using $\alpha_1 + \alpha_2 = 1-\gamma$, through a direct calculation we find that
    \begin{align*}
        g(M) - g(\id_{p\times p})%&= \frac{\sigma^2 \gamma}{2(1-\gamma)} (\lambda^2-1)a_1\cdot \bigbrace{  \frac{1}{ -a_1(\lambda^2-1)+(1-\gamma)\lambda^2 } - \frac{1}{a_1(\lambda^2-1) + (1-\gamma)}} \\
        &= \frac{\sigma^2 }{ n_1+n_2-p} \sum_{i=1}^{p/2} \frac{(\lambda_i^2-1)^2 \alpha_1 \left[ \alpha_1 - \alpha_2\right] }{[\alpha_1 + \lambda_i^2 \alpha_2][\lambda_i^2 \alpha_1 + \alpha_2]} .
    \end{align*}
    %in this example is equal to
    %$\frac{p}{2(n_1 + n_2)} f(\lambda)$, where
    %\[ f(\lambda) = {(\lambda^{-2} a_1 + a_2)^{-1} + (\lambda^2 a_1 + a_2)^{-1}}. \]
    %Using the fact that $a_1 + a_2 = 1 - \frac{p}{n_1 + n_2}$, we can verify
    %\begin{align*}
    %	f(\lambda) - f(1) &= \left(\lambda^2 a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
    %	&+ \left(\lambda^{-2} a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
    %	&- \frac{2(n_1 + n_2)}{n_1 + n_2 - p} \\
    %	&= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
    %\end{align*}
    %\begin{align*}
    %	f(\lambda) - f(1) &= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
    %\end{align*}
    %where $g(\lambda, a_1) \ge 0$.
    %and can be derived from algebraic calculations (details omitted).
    We claim that $\alpha_1 > \alpha_2$ if and only if $n_1 > n_2$, which then explains the dichotomy.
    In fact, if $\alpha_1 > \alpha_2$, then the first part of equation \eqref{eq_a12extra} gives that $\alpha_1 > (1-\gamma)/2$.
    The second part of equation \eqref{eq_a12extra} gives that
    \begin{align*}
     \frac{n_1}{n_1 + n_2} &> \alpha_1 + \frac{1}{n_1+n_2} \sum_{i=1}^{p/2}\left(\frac{\lambda_i^2}{\lambda_i^2+1}+\frac{\lambda_i^{-2}}{\lambda_i^{-2}+1}\right) = \frac{1-\gamma}{2}+\frac{\gamma}{2}=\frac{1}{2}.
    \end{align*}
    %\begin{align*}
    % \frac{n_1}{n_1 + n_2} &=a_1 + \frac1{n_1 + n_2}\cdot \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2}} \\
    %	&> a_1 + \frac{p}{2(n_1+n_2)} \left(\frac{\lambda^2}{\lambda^2+1}+\frac{\lambda^{-2}}{\lambda^{-2}+1}\right) =\frac{1}{2}.
    %\end{align*}
    This is equivalent to $n_1>n_2$. The other direction follows from a similar argument. % Similarly, if $a_1<a_2$, equations  \eqref{eq_a12extra000} and  \eqref{eq_a12extra} give that $a_1 < \frac{n_1 + n_2-p}{2 (n_1 + n_2)}$ and $n_1<n_2$. Thus, we conclude that $f(\lambda) \ge f(1)$ if and only if $n_1 \ge n_2$.
    %\end{example}
\end{proof}

Figure \ref{fig_sec3_covariate} illustrates a special case where $\lambda_1 = \cdots = \lambda_{\floor{p /2}} = \lambda > 0$ for $M^{\top}M$ and the rest of the eigenvalues are all equal to $1 / \lambda$.
Thus, $\lambda > 0$ captures the degree of covariate shift: higher $\lambda$ implies worse covariate shift.
We observe that our theoretical estimates using Corollary \ref{cor_hps_cov} matches the empirical risks incredibly accurately.
As a result, we indeed observe the dichotomy shown in Claim \ref{claim_dichotomy}.
Furthermore, for higher $\lambda$, HPS' excess risk decreases slower--indicating a worse rate of ``transfer'' from task one. 
As a remark, the classical work of \citet{david2010impossibility} have likewise shown impossibility results for transfer learning under covariate shift in a classification setting.
By contrast, our analysis applies to a (high-dimensional) regression setting.
\end{example}

\begin{example}
Our second example illustrates that the effect of covariate shift worsens as the sample size of task one increases.
We consider a set of covariate shift matrices $M^{\top} M$ whose determinant are all equal to one, and the eigenvalues are bounded between $\tau^2$ and $1 / \tau^2$.
Let $\cS$ denote the set of $M\in\real^{p\times p}$ such that $M^{\top} M$ satisfies both conditions.
\begin{claim}[Identical covariance provides approximately optimal transfer under imbalanced dataset sizes]\label{prop_covariate}
    Recall that $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
	We have the following result regarding $g(\id_{p \times p})$:
	\begin{align} g(\id_{p\times p}) \le \bigbrace{1+ {\frac{n_2}{\tau^2 n_1}  }} g(M), \text{ for any } M \in \cS. \label{eq_claim_id}
	\end{align}
\end{claim}
\begin{proof}
    We can write down the trace of $(\alpha_1 M^{\top} M + \alpha_2)^{-1}$ using the eigenvalues of $M^{\top} M$ as follows:
    \begin{align*}
        (\alpha_1 M^{\top} M + \alpha_2)^{-1} &= \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2} \\
        &\ge \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2 \cdot {\lambda_i^2} / {\tau^2}} \tag{since $\lambda_i \ge \tau$, for any $i$} \\
        &= \frac 1 {\alpha_1 + \alpha_2 / \tau^2} \sum_{i=1}^ p \frac 1 {\lambda_i^2} \\
        &\ge \frac 1 {\alpha_1 + \alpha_2 /\tau^2} p\cdot \bigbrace{\prod_{i=1}^p \frac 1 {\lambda_i^2}}^{1/ p} \tag{by the AM-GM inequality} \\
        &= \frac{p}{\alpha_1 + \alpha_2 / \tau^2} \tag{since $\prod_{i=1}^p \lambda_i^2 = \det(M^{\top} M) = 1$}
    \end{align*}
    Next, recall from equation \eqref{g_id} that
    $g(\id_{p\times p}) = \frac{\sigma^2 p}{n_1 + n_2 - p}$.
    Thus, equation \eqref{eq_claim_id} follows if we show that (after rearranging terms)
    \begin{align}
        \frac{n_1 + n_2}{n_1 + n_2 - p} \cdot \frac 1 {\alpha_1 + \frac{\alpha_2}{\tau^2}} \le 1 + \frac{n_2}{\tau^2 \cdot n_1}. \label{eq_claim_id_step}
    \end{align}
    From the second part of equation \eqref{eq_a12extra}, we have that
    \begin{align*}
        \frac{n_1} {n_1 + n_2} &= \alpha_1 + \frac 1 {n_1 + n_2} \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2\alpha_1}{\lambda_i^2\alpha_1 + \alpha_2}} \\
        &< \alpha_1 + \frac p {n_1 + n_2}.
    \end{align*}
    Thus, $\alpha_1 > \frac{n_1 - p}{n_1 + n_2}$, which implies that $\alpha_2 < \frac{n_2}{n_1 + n_2}$.
    Hence, $\alpha_1 + \frac{\alpha_2}{\tau^2} = 1 - \frac{p}{n_1 + n_2} - \alpha_2 + \frac{\alpha_2}{\tau^2} \le 1 - \frac{p}{n_1 + n_2} + \frac{n_2}{n_1 + n_2} \frac 1 {\tau^2}$, which implies
    equation \eqref{eq_claim_id_step} by straightforward calculations.
    Thus, the proof is complete.
\end{proof}
As a corollary of Claim \ref{prop_covariate}, when $n_1$ is much larger than $n_2 / \tau^2$, the excess risk of HPS is approximately optimal in the set of all possible $M\in\cS$ when there is no covariate shift between dataset one and two.
\end{example}
